---
title: "Lab: Logistic Regression"
subtitle: "Princeton University"
author: "Younes Strittmatter"
output: 
  tufte::tufte_html:
    css: 
    tufte_variant: "envisioned"
    highlight: github-dark
    fig_height: 10
    fig_width: 16
    toc: true
    toc_depth: 1
execute: 
  message: false
  warning: false
format: 
  html:
    code-fold: true
    code-overflow: wrap
engine: knitr
---

## Data: General Social Survey

The General Social Survey (GSS) has been used to measure trends in attitudes and behaviors in American society since 1972. In addition to collecting demographic information, the survey includes questions used to gauge attitudes about government spending priorities, confidence in institutions, lifestyle, and many other topics. A full description of the survey may be found here.

The data for this lab are from the 2016 General Social Survey. The original data set contains 2867 observations and 935 variables. We will use and abbreviated data set that includes the following variables:

`natmass`: Respondent’s answer to the following prompt:

“We are faced with many problems in this country, none of which can be solved easily or inexpensively. I’m going to name some of these problems, and for each one I’d like you to tell me whether you think we’re spending too much money on it, too little money, or about the right amount…are we spending too much, too little, or about the right amount on mass transportation?”

`age`: Age in years.

`sex`: Sex recorded as male or female

`sei10`: Socioeconomic index from 0 to 100

`region`: Region where interview took place

`polviews`: Respondent’s answer to the following prompt:

“We hear a lot of talk these days about liberals and conservatives. I’m going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal - point 1 - to extremely conservative - point 7. Where would you place yourself on this scale?”

The data are in gss2016.csv in the data folder.

## EDA

> Let’s begin by making a binary variable for respondents’ views on spending on mass transportation. Create a new variable that is equal to “1” if a respondent said spending on mass transportation is about right and “0” otherwise. Then plot the proportion of the response variable, using informative labels for each category.

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(modelsummary)
library(tidyr)
library(knitr)
library(easystats)
library(broom)
library(emmeans)
library(marginaleffects)
library(performance)
library(arm)
library(modelsummary)
library(here)
```

```{r}
# Set the working directory
setwd(here::here("pages/logistic-regression"))
# load data
data <- read.csv("gss2016.csv")
# look at the data 
head(data)
# See what `natmass` levels exist
unique(data$natmass)
```

Fill in the "\_\_\_\_" below to encode the binary variable

```{r}
data <- data %>%
   mutate(mass_trans_spend_right = as.numeric(natmass == "About right"))
head(data$mass_trans_spend_right)
```


```{r}
#Get proportions
mass_spend_summary <- data %>%
  count(mass_trans_spend_right) %>%
  mutate(proportion = n / sum(n))

#Look at the dataframe structure. And make sure it's in a format that you can use for plotting.
#Change structure if neederd
mass_spend_long <- mass_spend_summary %>% mutate(opinion="Proportion")

#Factorise for plot
mass_spend_long$mass_trans_spend_right <- as.factor(mass_spend_long$mass_trans_spend_right)

#Make plot
#Hint: geom_bar lets you make stacked bar charts
ggplot(mass_spend_long, aes(x = opinion, y = proportion, fill = mass_trans_spend_right)) + geom_bar(stat = "identity")

```

> Recode `polviews` so it is a factor with levels that are in an order that is consistent with question on the survey. Note how the categories are spelled in the data.

```{r}

data <- data %>%
  mutate(polviews = factor(polviews,
                           levels = c("Extremely liberal", "Liberal", "Slightly liberal",
                                      "Moderate", "Slghtly conservative","Conservative", "Extrmly conservative"),
                           ordered = TRUE))
```

> Make a plot of the distribution of `polviews`

```{r}
# Get proportions
poly_summary <- data %>%
  count(polviews) %>%
  mutate(proportion_o = n / sum(n))

# Keep the data structure appropriate for plotting
poly_long <- poly_summary %>% mutate(opinion="Proportion")

# Convert polviews into a factor if not already
poly_long$polviews <- as.factor(poly_long$polviews)

# Plot the proportions
ggplot(poly_long, aes(x = opinion, y = proportion_o, fill = polviews)) + 
  geom_bar(stat = "identity")
```

> Which political view occurs most frequently in this data set?
>
> \_\_\_\_\_

> Make a plot displaying the relationship between satisfaction with mass transportation spending and political views. Use the plot to describe the relationship the two variables.

```{r}
# Plot the relationship between satisfaction with mass transportation spending and political views
ggplot(data, aes(x = polviews, fill = mass_trans_spend_right)) +
  geom_bar(position = "fill") +
  labs(title = "Relationship between Political Views and Satisfaction with Mass Transportation Spending",
       x = "Political Views",
       y = "Proportion",
       fill = "Satisfaction with Spending") +
  theme_minimal()
```

The more conservative one's political views are the more they think the amount of spending on mass transportation is correct.

-   We’d like to use age as a quantitative variable in your model; however, it is currently a character data type because some observations are coded as "89 or older".

> Recode age so that is a numeric variable. Note: Before making the variable numeric, you will need to replace the values "89 or older" with a single value.

```{r}
data <- data %>%
  mutate(age = if_else(age == "89 or older", "89", age), 
         age = as.numeric(age))
```

> Plot the frequency distribution of age.

```{r}
ggplot(data, aes(x = age)) +
  geom_histogram(binwidth = 1,color='black', alpha = 0.7) +
  labs(title = "Age Distribution",
       x = "Age",
       y = "Frequency") +
  theme_minimal()
```

## Logistic regression

> Let’s start by fitting a logistic regression model with just the intercept

```{r}
intercept_only_model <- glm(
  mass_trans_spend_right ~ 1, 
  data = data, 
  family = binomial
) 

intercept_only_model %>% 
  tidy() %>%
  kable()

```

> Interpret the intercept in the context of the data. You can do this by converting the $\beta_0$ parameter out of the log-odds metric to the probability metric. Make sure to include the 95% confidence intervals. Then interpret the results in a sentence or two--what is the basic thing this probability tells us about?

```{r}
b0 <- coef(intercept_only_model)[1] # get coef

b0_transformed <- exp(b0) / (1 + exp(b0)) # logistic transform

ci_lower = b0 - 1.96 * 0.0393685
ci_upper = b0 + 1.96 * 0.0393685

#transforming confidence intervals of coefficients into probabilities
p_lower = exp(ci_lower) / (1 + exp(ci_lower))
p_upper = exp(ci_upper) / (1 + exp(ci_upper))

b0_transformed
p_lower
p_upper

```

Interpretation: The intercept-only model's intercept (β0) represents the log-odds of a respondent being satisfied with mass transportation spending when all predictors are at their reference levels. Converting this to the probability metric, the probability of a respondent being satisfied with mass transportation spending is approximately 0.53 (95% CI [0.51, 0.55]). This tells us that, on average, about 53% of respondents are satisfied with the current spending on mass transportation.

> Now let's fit a model using the demographic factors - `age`,`sex`, `sei10` - to predict the odds a person is satisfied with spending on mass transportation. Make any necessary adjustments to the variables so the intercept will have a meaningful interpretation. *Neatly display the model coefficients (do not display the summary output)*

```{r}
#make sure that sex is a factor (i.e. to make sure R knows it's binary/categorical, and not continuous)
data <- data %>%
  mutate(sex = factor(sex, levels = c("Male", "Female")))

#fit with glm()
m1 <- glm(mass_trans_spend_right ~ age + sex + sei10, data = data, family = binomial)

#produce tidy output of model coefficients
m1 %>% 
  tidy() %>%
  kable()
```

> Consider the relationship between sex and one’s opinion about spending on mass transportation. Interpret the coefficient of sex in terms of the logs odds and OR of being satisfied with spending on mass transportation. What are the predicted probabilities for males and females on support for spending on mass transportation? Please include the 95% CIs around each estimate.

```{r}
m1 %>% 
  tidy() %>%
  kable()

m1 %>% 
  tidy(exponentiate = TRUE) %>%
  kable()

bsex <- coef(m1)["sexFemale"]

ci_lower_lo = bsex - 1.96 * 0.0798020
ci_upper_lo = bsex + 1.96 * 0.0798020

ci_lower_or = 1.29 - 1.96 * 0.0798020
ci_upper_or = 1.29 + 1.96 * 0.0798020

emm_sex <- emmeans(m1, "sex", type = "response")

ci

```

If you did this right, you'll find that being female (as compared to male) is associated with an increase in the log-odds of being satisfied with spending on mass transportation by 0.2557439 units (95% CI \[0.09, 0.41\]), holding all other variables constant. This equates to the odds of thinking the spending amount is right in females being 1.29 times the odds of thinking this in men (95% CI \[1.13, 1.44\]).

The predicted probability for females to be satisfied with spending on mass transportation is 55.9% (95% CI \[53.3%, 58.5%\]) and that of males is 49.5% (95% CI \[46.7%, 52.4%\]).\
\
Verify this.

> Next, consider the relationship between age and one’s opinion about spending on mass transportation. Interpret the coefficient of age in terms of the logs odds and OR of being satisfied with spending on mass transportation. Please include the 95% CIs around each estimate.

```{r}
bage <- coef(m1)["age"]

ci_lower_age = bage - 1.96 * 0.005
ci_upper_age = bage + 1.96 * 0.005

#transforming confidence intervals of coefficients into odds ratios
or_age = exp(bage)
or_lower_age = exp(ci_lower_age)
or_upper_age = exp(ci_upper_age)

bage
or_age
or_lower_age
or_upper_age
```

A one unit increase in age is associated with a decrease in the log-odds of being satisfied with spending on mass transportation by `r bage` (95% CI [`r ci_lower_age`, `r ci_upper_age`]), holding all other variables constant. The odds ratio is `r or_age` (95% CI [`r or_lower_age`, `r or_upper_age`])which confirms the negative relationship implied by the log-odds coefficient. Specifically, for each additional unit of age, the odds of being satisfied with mass transportation spending decrease by a factor of about `r or_age`, or approximately `r 100 - or_age * 100`% per unit increase in age, holding other factors constant.

> Consider the relationship between SES and one’s opinion about spending on mass transportation. Interpret the coefficient of SES in terms of the logs odds and OR of being satisfied with spending on mass transportation. Please include the 95% CIs around each estimate. ß
```{r}
bses <- coef(m1)["sei10"]

ci_lower_ses = bses - 1.96 * 0.0023
ci_upper_ses = bses + 1.96 * 0.0023

#transforming confidence intervals of coefficients into odds ratios
or_ses = exp(bses)
or_lower_ses = exp(ci_lower_ses)
or_upper_ses = exp(ci_upper_ses)

bses
or_ses
or_lower_ses
or_upper_ses
```

A one unit increase in SES index is associated with a decrease in the log-odds of being satisfied with spending on mass transportation by 0.0062 units (95% CI \[-0.0107, -0.0017\]), holding all other variables constant. The odds ratio is less than 1 (0.9937922), which confirms the negative relationship implied by the log-odds coefficient. Specifically, for each additional unit of SES index, the odds of being satisfied with mass transportation spending decrease by a factor of about 0.993, or approximately 0.7% per unit increase in SES index, holding other factors constant (95% CI \[0.989, 0.998\]).

## Marginal effects

-   Let's examine the results on the probability scale.

> Calculate the marginal effects of sex, age, and SES on mass transportation spending. You can use the `margins` package function `margins` discussed in your textbook or you can use the `marginaleffects` package `avg_slope` `avg_comparisons` discussed in lecture. Interpret each estimate.

```{r}
# Calculate the marginal effects using the marginaleffects package
marginal_effects <- avg_comparisons(m1, comparison = "difference")

# Display the results
marginal_effects %>% 
  kable()

# Extract and interpret the marginal effects
marginal_effects_summary <- marginal_effects %>%
  summarise(
    term = term,
    estimate = estimate,
    std.error = std.error,
    conf.low = conf.low,
    conf.high = conf.high,
    p.value = p.value
  )


marginal_effects_summary

age_estimate <- marginal_effects_summary$estimate[marginal_effects_summary$term == "age"]
age_conf_h <- marginal_effects_summary$conf.high[marginal_effects_summary$term == "age"]
age_conf_l <- marginal_effects_summary$conf.high[marginal_effects_summary$term == "age"]
age_p <- marginal_effects_summary$p.value[marginal_effects_summary$term == 'age']

ses_estimate <- marginal_effects_summary$estimate[marginal_effects_summary$term == "sei10"]
ses_conf_h <- marginal_effects_summary$conf.high[marginal_effects_summary$term == "sei10"]
ses_conf_l <- marginal_effects_summary$conf.high[marginal_effects_summary$term == "sei10"]
ses_p <- marginal_effects_summary$p.value[marginal_effects_summary$term == 'sei10']

fem_estimate <- marginal_effects_summary$estimate[marginal_effects_summary$term == "sex"]
fem_conf_h <- marginal_effects_summary$conf.high[marginal_effects_summary$term == "sex"]
fem_conf_l <- marginal_effects_summary$conf.high[marginal_effects_summary$term == "sex"]
fem_p <- marginal_effects_summary$p.value[marginal_effects_summary$term == 'sex']



```

-   The marginal effect of age is `r age_estimate` (95% CI \[`r age_conf_l`, `r age_conf_h`\]). So, for each additional unit increase of age, the probability of being satisfied with mass transportation spending decreases by approximately `r age_estimate*100` percentage points, holding other factors constant (p = `r age_p`).

-   The marginal effect of SES is `r ses_estimate` (95% CI \[`r ses_conf_l`, `r ses_conf_h`\]). For each one-unit increase in the socioeconomic index, the probability of being satisfied with mass transportation spending decreases by approximately `r ses_estimate * 100` percentage points, holding other variables constant (p = `r ses_p`).

-   The marginal effect for being female compared to male is `r fem_estimate` (95% CI \[`r fem_conf_l`, `r fem_conf_h`\]). This indicates that females are, on average, about `r fem_estimate * 100` percentage points more likely than males to be satisfied with mass transportation spending, holding other factors constant (p = `r fem_p`).

## Model comparison

-   Now let’s see whether a person’s political views has a significant impact on their odds of being satisfied with spending on mass transportation, after accounting for the demographic factors.

> Conduct a drop-in-deviance/likelihood ratio test to determine if polviews is a significant predictor of attitude towards spending on mass transportation. Name these two models `fit2` and `fit3`, respectively. Compare the two models.

```{r}

fit2 <- glm(mass_trans_spend_right ~ age + sex + sei10, data = data, family = binomial)

fit3 <- glm(mass_trans_spend_right ~ age + sex + sei10 + polviews, data = data, family = binomial)

test_likelihoodratio(fit2, fit3) %>% kable()

```

> Is the model with polviews better than the model without?

-   Yes.

## Visualization

-   Let’s plot the results

-   We next use the model to produce visualizations:

  -   Given the code below, interpret what is being plotted:

    -   pol_plot : This plot shows the predicted probability of being satisfied with mass transportation spending across different political views, with error bars representing the 95% confidence intervals.

    -   sex_plot : This plot shows the predicted probability of being satisfied with mass transportation spending for males and females, with error bars representing the 95% confidence intervals.

    -   ses_plot: This plot shows the predicted probability of being satisfied with mass transportation spending across different levels of socioeconomic status (SES), with a confidence interval band representing the 95% confidence intervals.

  ::: callout-tip
  -   adjust the various settings in your plot to make it look professional.

  -   You can use `ggeffects` to get the predicted probabilities for these models.
  :::

```{r}
#| fig-width: 12
#| fig-height: 10
library(ggeffects)


colors <- c("Extremely liberal" = "black",
            "Liberal" = "#0e2f44",  # Dark blue
            "Slightly liberal" = "#1d5a6c",  # Less dark blue
            "Moderate" = "#358ca3",  # Medium blue
            "Slghtly conservative" = "#71b9d1",  # Light blue
            "Conservative" = "#a6dcef",  # Lighter blue
            "Extrmly conservative" = "#d0f0fd")  # Very light blue

pp_pol <- ggemmeans(fit3, terms = c("polviews"))

# Adjusted plot with gradient colors
pol_plot <- ggplot(pp_pol, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  scale_color_manual(values = colors) +
  labs(title = "Effect of Political Views on Satisfaction with Mass Transportation",
       x = "Political Views", y = "Predicted Probability",
       color = "Political Views") +
  theme_minimal()

pol_plot

pp_sex <- ggemmeans(fit3, terms = c("sex"))

sex_plot <- ggplot(pp_sex, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(title = "Effect of Sex on Satisfaction with Mass Transportation",
       x = "Sex", y = "Predicted Probability",
       color = "Sex") +
  theme_minimal()

pp_sex

pp_ses <- ggemmeans(fit3, terms = "sei10")


ses_plot <-  ggplot(pp_ses, aes(x = x, y = predicted)) +
  geom_line(color = "#2c7fb8", size = 1) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#2c7fb8", alpha = 0.2) +  # Add a confidence interval band
  labs(title = "Effect of SES on Satisfaction with Mass Transportation",
       x = "Socioeconomic Status", y = "Predicted Probability") +
  theme_minimal() +
  theme(legend.position = "none")  
ses_plot

```

## Model Assumptions

-   Is the logistic model a good choice for this data?

```{r}
binned_residuals(fit2)
```

::: callout-note
Answer: The model is not a very good choice since it doesn't fit the data well. Only 86% of the residuals are in the erro bounds while >95% are expected.
:::

## Model fit

-   Calculate the $R^2$ for this model

```{r}

r2 <- r2_mcfadden(fit2)

R2 = r2$R2
```

> R2 interpretation: The McFadden's R2 value for this model is an indicator of the model's goodness of fit. A higher R2 value suggests a better fit, meaning the model explains a larger proportion of the variance in the response variable. Here only `r R2 * 100`% of the variance are explained suggesting a low fit.
>
> Next, Take a look at the binned residual plots for each continuous predictor variable and look at linearity. Is there a predictor that sticks out? What can we do to improve model fit in this case?

```{r}
binned_residuals(fit2, term="sei10")
binned_residuals(fit2, term="age")

binned_residuals(fit2, term="sei10") %>% plot(show_dots=TRUE)
binned_residuals(fit2, term="age") %>% plot(show_dots=TRUE)

```

::: callout-note
-   ::: callout-note
    Age sticks out in a positive way.
    :::
:::

## Testing Polviews

```{r}
emmeans(fit3, "polviews") %>% pairs() %>% as.data.frame() %>% filter(p.value < .05)

emmeans(fit3, "polviews", type="response") %>% pairs() %>% as.data.frame() %>% filter(p.value < .05)
```

-   Conservatives are `r 1/.37` times more likely to support mass transit spending compared to extremely liberal and liberal

-   Extreme liberals are .37, , .4, .43 times as likely to support spending compared to conservatives, moderates and slight conservatives

-   Extreme conservatives are `r 1/.33`, `r 1/.48` times more likely to support mass spending than liberals and slight liberals

-   Liberals are .49, .46 times more likely to support spending than moderates and slight conservatives.

## Conclusion

The model best predicts satisfaction with mass transportation when it includes age, sex, socioeconomic status (SES), and political views (polviews) as predictors. Specifically, the results indicate that:

- Conservatives tend to express greater satisfaction with mass transportation compared to liberals.
- Females report higher satisfaction with mass transportation than males.
- Individuals with a lower socioeconomic status are more satisfied with mass transportation than those with a higher socioeconomic status.

These findings suggest that political ideology, gender, and socioeconomic background all play a role in shaping public attitudes toward mass transportation.

```{r model-summary, results='asis', echo=FALSE , fig.cap="Table 1"}

fit_anova = anova(fit3, test="Chisq")

fit_anova %>%
  kable()


```

```{r plot-sexviews, fig.cap="Figure 1: Effect of Sex on Satisfaction with Mass Transportation", echo=FALSE}
sex_plot
```

```{r plot-sesviews, fig.cap="Figure 2: Effect of SES  on Satisfaction with Mass Transportation", echo=FALSE}
ses_plot
```

```{r plot-polviews, fig.cap="Figure 3: Effect of Political Views on Satisfaction with Mass Transportation", echo=FALSE}
pol_plot
```
